### 2020-06-18 paper/sex_selection_spacing-ver4.tex
Add three main questions and define sex selection

\date{February 2020}

@@

However, the males-to-females ratio at birth also increased dramatically over this period
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.
Chorionic villus sampling (CVS), amniocentesis, and ultrasound all allow parents to learn 
a fetus' sex, and access to these technologies expanded rapidly.
With a continued strong preference for sons, the result has been an increase in sex 
selection, the selective abortion of female fetuses based on prenatal sex determination.%
\footnote{
CVS requires the shortest period of gestation---eight to twelve
weeks---which means that abortions can occur in the first trimester.
Amniocentesis can be performed after fourteen weeks but requires three
to four weeks before the result is available.
Hence, any abortion would not take place until more than halfway through the second 
trimester.
Ultrasound has the advantages of being non-invasive and relatively cheap.
The sex of a fetus can be determined in the third month of gestation if
it is a boy and the fourth month if it a girl.
}

The main question I examine is: how did birth spacing respond to these
significant changes, especially the introduction and spread of sex selection? 
This question is motivated by two factors. 
First, prior research has failed to appreciate that each abortion increases the
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover; 
otherwise, the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
The second part is the waiting time to conception, which is between one and six months. 
Finally, sex determination tests are reliable only from three months of gestation onwards.
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
but we know little about how much. 
Second, more female education, higher household income, and low and declining female labor 
force participation are all likely to impact birth spacing, and the combined changes in 
birth spacing may, therefore, be much more substantial in India than what we have observed 
in other countries.

Understanding birth spacing is valuable not only because it contributes not only to our 
knowledge about fertility decisions but also because of the potential effects of birth 
spacing on outcomes such as fertility measures and mortality. 
I focus on two questions.

First, to what extent do changes in birth spacing bias our standard fertility estimates?
The tempo effect that arises from longer birth intervals means that the total fertility 
rate is a downward biased estimate of cohort fertility 
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
Hence, if birth intervals increased substantially over time, India's fertility may 
be higher and have declined slower than generally accepted.

Second, how do the changes in birth spacing and the increased use of abortions impact 
children's survival chances?
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer birth spacing then potentially lowers mortality and morbidity risk 
\citep{Conde-Agudelo2012,Molitoris2019}.
There may, however, be counteracting effects if the longer spacing arises because of 
multiple abortions.
There has, so far, been no attempt to examine how infant mortality has responded 
to the increase in sex selection.

To examine how birth spacing in India has changed over time and across groups with the 
introduction of sex selection, I apply a competing risk hazard model to the birth 
histories of Hindu women, covering the period 1972 to 2016, using data from the four 
National Family and Health Surveys (NFHS).
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son 
but also works in cases without prenatal sex selection.
The empirical model allows me to predict fertility, taking into
account both the likelihood of parity progression and the likelihood of having a child of 
a specific sex.
I use the same data to examine how infant mortality risk has changed with the changing 
birth spacing and whether sex selection affects infant mortality.

% [Hypotheses / questions]
There are four main sets of results.

First, birth intervals have increased for all education groups over the four decades, 
the more so, the higher the parity and the higher the education level.
Women who are most likely to use sex selection---well-educated women with no
sons---have seen the most substantial increases in birth intervals and the most 
biased sex ratios.
As a result, we now see cases that reverse the traditional spacing
pattern, with women with no sons having longer birth intervals than
women with sons.
Those least likely to use sex selection, women with less education in rural areas, 
still follow the traditional pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
Women with no sons still are the most likely to have an additional birth.
Although the probability of having a next child has declined for this group over time, the 
declines are often substantially smaller than if a boy is present.

Second, there is little difference over time and across groups in the likelihood of 
having a very short birth interval given the sex composition of prior children, 
especially for the lower parities.
Hence, the increases in average spacing come predominantly from the longer birth intervals
getting even longer over time, driven in no small degree by substantial use of sex 
selection.
The exception to this result is that the better-educated women with a \emph{very high} 
predicted use of sex selection appear to conceive later precisely because the next child 
is almost certainly a boy.

Third, the fertility rate substantially overestimated how fast cohort fertility fell 
in the 1990s and early 2000s when spacing began to increase.
Although predicted cohort fertility and the fertility rate have lately begun to converge,
the predicted cohort fertility is still ten to twenty percent higher than the
fertility rate.
Furthermore, even with the convergence, predicted cohort fertility is still at or above 
replacement level, except for the best-educated urban women.

Finally, infant mortality risk has declined substantially over time for all groups,
but fastest for the lower education groups, who are now close to the level of women with 
the most education.
The mortality risk is, however, still inversely related to education level, especially for 
very short birth intervals.
There is no evidence that the increasing use of sex selection is associated with higher 
mortality risk.

>>>>

\date{June 2020}

@@

However, India also witnessed the appearance of sex selection---the selective 
abortion of female fetuses based on prenatal sex determination. 
Sex selection became available in the mid-eighties with the introduction of technologies 
such as ultrasound and amniocentesis. 
Combined with continued strong son preference, the result was a dramatic increase in the 
male-to-female ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.

The main question I examine is: how did birth spacing respond to these
changes, especially the introduction and spread of sex selection? 
The motivation is twofold.
First, prior research has failed to appreciate that each abortion increases the
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover; 
otherwise, the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is between one and six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
but we know little about how much. 
Second, more female education, higher household income, and low and declining female labor 
force participation are all also likely to impact birth spacing, and the combined changes 
may be much more substantial in India than what we have observed in other countries.

Understanding birth spacing is valuable because it contributes to our knowledge about 
fertility decisions and because of birth spacing's potential effects on outcomes such as 
fertility measures and mortality. 
I focus on two questions.

First, to what extent do changes in birth spacing bias our standard fertility estimates?
The tempo effect that arises from longer birth intervals means that the total fertility 
rate is a downward biased estimate of cohort fertility 
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
Hence, if birth intervals increased substantially over time, India's fertility may 
be higher than generally accepted.

Second, how do the changes in birth spacing and the increased use of abortions impact 
children's chances of survival?
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer birth spacing then potentially lowers mortality and morbidity risk 
\citep{Conde-Agudelo2012,Molitoris2019}.
There may, however, be counteracting effects if the longer spacing arises because of 
multiple abortions.
There has, so far, been no attempt to examine how infant mortality has responded 
to the increase in sex selection.

To examine how birth spacing has changed over time and across groups with the introduction 
of sex selection, I apply a competing risk hazard model to the birth histories of Hindu 
women, covering the period 1972 to 2016, using data from the four National Family and 
Health Surveys (NFHS).
The model directly incorporates the effects of sex-selective abortions on birth intervals
and the likelihood of a son, but it also works with no sex selection.
The empirical model allows me to predict fertility, taking into
account both the likelihood of parity progression and the likelihood of having a child of 
a specific sex.
I use the same data to examine how infant mortality risk has changed with the changing 
birth spacing and whether sex selection affects infant mortality.

% [Hypotheses / questions]
There are four main results.

First, birth intervals have increased for all education groups over the four decades, but 
more so, the higher the parity and the higher the education. 
Women most likely to use sex selection---well-educated women with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with women 
with no sons having \emph{longer} birth intervals than women with sons. 
Those least likely to use sex selection---women with less education in rural areas---still 
follow the traditional pattern of short spacing when they have girls and little evidence 
of sex selection.

Second, there is little change in the likelihood of having a very short birth interval. 
Hence, the increased average spacing comes mainly from the longer birth intervals getting 
even longer, primarily driven by substantial use of sex selection. 
The exception to this result is that those better-educated women who have a very high 
predicted use of sex selection conceive later precisely because the next child is almost 
certainly a boy.

Third, the fertility rate substantially overestimated how fast cohort fertility fell 
in the 1990s and early 2000s when spacing began to increase.
Although predicted cohort fertility and the fertility rate have lately begun to converge,
the predicted cohort fertility is still ten to twenty percent higher than the
fertility rate.
Furthermore, even with the convergence, predicted cohort fertility is still at or above 
replacement level for all but the best-educated urban women.

Finally, infant mortality risk has declined substantially over time for all groups,
but fastest for the lower education groups, who are now close to the level of women with 
the most education.
However, the mortality risk is still inversely related to education level, especially for 
very short birth intervals.
There is no evidence that the increasing use of sex selection is associated with higher 
mortality risk.
